---
title: "Zoning For Jobs: Industrial Land-Use Preservation\n in the nation's largest cities"
author: "Jamaal Green"
date: "October 9, 2017"
output: 
  beamer_presentation:
    incremental: TRUE
    theme:  "Boadilla"
    slide_level:  2
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(cache = TRUE, tidy = FALSE)

if(!require(pacman)){install.packages("pacman"); library(pacman)}
p_load(readr, tidyr, dplyr, ggplot2)
```

#Introduction

##Some Observations

- Many urban labor markets are recovering/have recovered from the Great Recession
- This growth has not translated to greater wages for lesser skilled workers
- This growth is placing generational levels of demand on relatively centrally located land

##So What?

- Absent direct regulatory steps (increasing min. wage, mandating paid leave) planners are left with human capital (skills upgrading etc.) or *industrial* strategies
- Industrial jobs remain, on average, better jobs for workers with less formal education than equivalent service jobs therefore planners should work on supporting/growing such jobs
- Growing industrial jobs requires planners to make or *preserve* space for industrial users


##Industrial Land Matters (again)

- Leigh and Hoelzel (2012)
    + "Smart Growth's Blindside"...planning practice for the past 30 years has been one of explicitly chasing industrial users out of the city
- Chapple (2014)
    + Available industrial land encourages job growth
- Lester, Kaza, Kirk (2013)
    + Converting industrial land to non-industrial has mixed labor market outcomes we should consider
- Leigh et al (2014)
    + PAS *Sustainable Urban Industrial Development*

##But...

What cities are protecting industrial land? What policies/techniques are they using? 

#Methods

##My approach...

Content analysis of the zoning codes and most recent comp plans of the nation's 50 largest cities

##Identifying the Codes

> - Discursive
  + Frames industrial land as an asset and not as a nuisance to be manage
> - Functional
  + Distinct and protective- a unique code/layer that adds some form of additional protection and coding what the additional restrictions are (with particular focus on increased residential and commercial restrictions)

##Baltimore's Marine Industrial Districts

"THE MI MARITIME INDUSTRIAL ZONING DISTRICT IS INTENDED TO ENSURE THE PRESERVATION OF LIMITED DEEP-WATER FRONTAGE OF THE PORT OF BALTIMORE FOR MARITIME USE BY DELINEATING AN AREA WHERE MARITIME SHIPPING CAN BE CONDUCTED WITHOUT THE INTRUSION OF NON-INDUSTRIAL USES AND WHERE INVESTMENT IN MARITIME INFRASTRUCTURE IS ENCOURAGED"

##Who Protects Their Industrial Land

| City          | State | Population, 2012 | Year Ordinance Passed |
|:---------------|:-----:| ----------------:| ---------------------:|
| Los Angeles   |   CA  |     3,857,799    |          2008         |
| Chicago       |   IL  |     2,714,856    |          1988         |
| San Diego     |   CA  |     1,338,348    |          2008         |
| Jacksonville  |   FL  |      836,507     |          2007         |
| San Francisco |   CA  |      825,863     |          2008         |
| Boston        |   MA  |      636,479     |          1978         |
| Seattle       |   WA  |      634,535     |          2002         |
| Baltimore     |   MD  |      621,342     |          2004         |
| Portland      |   OR  |      603,106     |          1988         |
| San Jose      |   CA  |      982,765     |          2007         |
| Minneapolis   |   MN  |      392,880     |          2009         |
| New York      |   NY  |     8,199,221    |          2006         |

##Cities Use Combination of Restrictions and Process

```{r, echo=FALSE, message=FALSE, warning=FALSE}

land_matrix <- read_csv("policy_matrix.csv")

land_long <- gather(land_matrix, key = "Policy", value = "Restriction", 2:9)

land_long$Policy <- factor(land_long$Policy, levels = c("Residential", "Commercial", "Educational", 
                                                        "City.Review", "Region.State",
                                                        "Industrial.Impact", "CBA", "Industrial.Support"))

land_long <- land_long %>% group_by(City) %>% mutate(PoliScore = sum(Restriction))
land_long <- land_long %>% ungroup() %>% mutate(Process = 
                                                  ifelse(Policy == "Residential" | 
                                                           Policy == "Commercial" |
                                                           Policy == "Educational", "Use",
                                                         "Process"))


p2 <- ggplot(land_long, aes(x = reorder(City, PoliScore), y = Policy, fill = Restriction)) + 
  geom_tile(color = "white") + 
  scale_fill_gradient(low = "white", high = "steelblue") + coord_flip() + 
  theme(axis.text.x  = element_text(angle = 45, hjust = 1, face = "bold", size = 14),
        axis.text.y = element_text(face = "bold", size = 14),
        axis.title.x = element_text(size = 16),
        axis.title.y = element_text(size = 16),
        legend.position = "none") + labs(x = "City")

p2
```


##Baltimore Maritime Industrial Districts

> - New zoning code
> - Passed in 2004 and strongest of the policies
> - Outright forbids nearly any non-industrial marine uses 
> - Rezoning applications require support of surrounding land owners, the city planning department, state port authority must be informed and give guidance

##Los Angeles Employment Protection Districts

> - A comprehensive plan overlay
> - Passed in 2008 after extensive industrial land use research period
> - Staff directed to reject residential rezones and to develop supportive policy
> - Industrial impact statements are required and notice to surrounding property owners
> - In less intensive zones, industrial conversion requires community benefits including relocation assistance and job training

##New York Industrial Business Zones

> - Local community-economic development organizations oversee them
> - Primarily zones for subsidy delivery including: tax relocation credits, job training credits, and small business support provided by the IBZ supervisor
> - Verbal statement from the Mayor to prevent residential conversion and hotel encroachment
> - Ongoing battle to include real restrictions in the city zoning code

#Conclusions

##Cities Care About Industrial Land

- Nearly a quarter of the nation's largest cities have some form of industrial land preservation policy
- Most of these policies are *new* and in flux
- These policies are *variable* ranging from outright bans to no net-loss strategies to verbal committments from city power structures
- The policies are primarily regulatory in nature and do not proactively support industrial users
- Indicative of a burgeoning set of industrial land *strategies* and links job creation more directly land-use

##Next Steps

- To what extent are these policies effective in protecting (and even growing) industrial jobs?
- What is the role of the planner in cities where these protections are in place? Do we see adequate enforcement? Are these policies reflective of actual city priorities and goals? 
