---
title: "Zoning For Jobs: Industrial Land-Use Policy\n in the nation's largest cities"
author: "Jamaal Green"
date: "October 4, 2017"
output: 
  beamer_presentation:
    incremental: TRUE
    theme:  "Boadilla"
    slide_level:  2
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(cache = TRUE, tidy = FALSE)

if(!require(pacman)){install.packages("pacman"); library(pacman)}
p_load(readr, tidyr, dplyr, ggplot2)
```

#Introduction and Background

##Uneven and Unequal Growth
```{r, message=FALSE, warning=FALSE, include=FALSE}
library(knitr)
```

- Many cities continue to grow steadily...
- BUT...the benefits of growth are ever more unequally distributed
- Housing gets (necessary) attention but inequality is also reflected in the labor market

##Good Jobs

- Many labor markets have recovered and are growing
- BUT...the *quality* of jobs available for workers of different educational backgrounds diverges
- Workers w/out degrees face flat wages, paltry benefits and few, if any, opportunities for advancement

##It's the Industry...

Absent major changes to labor market (Fight for $15!) planners are looking at *industrial* strategies

##Manufacturing Jobs are Still Good Jobs (mainly)

> - Mfg. workers w/out college degree still earn more (though narrowing) than non-mfg. workers
> - Mfg workers w/out college degree have access to better benefits than non-mfg workers

##Manufacturing is Urban

- The vast majority of mfg employment is in large metropolitan areas
- Demands for "advanced manufacturing" require extensive cooperation between producing and testing
- Consumer goods (particularly food) benefit from proximity to customers

##"Smart Growth's Blindside"

Planning for the past 30 years has focused on converting industrial land (and workers)

> - Industrial work is disappearing/obsolete
> - Industrial work/land is seen as dirty or non-sustainable
> - Industrial land is a relative fiscal loser 
> - People just prefer "mixed use"

##Protecting Industrial Land

Many cities realize industrial land is an *asset* and should be protected from conversion...but what cities are actually protecting their industrial land and how are they doing so?

#Methods

##Sample Cities

Reviewed the zoning codes and most recent comp plans of the nation's 50 largest cities (based on 2012 population)

##Code Identification

> - Distinct and protective- a *distinct* code from existing zoning or land designation and includes some stronger set of protections or rules
> - Industrial land as asset- the language of the code frames industrial land as an asset and not as a nuisance

##Who Protects Their Industrial Land

| City          | State | Population, 2012 | Year Ordinance Passed |
|:---------------|:-----:| ----------------:| ---------------------:|
| Los Angeles   |   CA  |     3,857,799    |          2008         |
| Chicago       |   IL  |     2,714,856    |          1988         |
| San Diego     |   CA  |     1,338,348    |          2008         |
| Jacksonville  |   FL  |      836,507     |          2007         |
| San Francisco |   CA  |      825,863     |          2008         |
| Boston        |   MA  |      636,479     |          1978         |
| Seattle       |   WA  |      634,535     |          2002         |
| Baltimore     |   MD  |      621,342     |          2004         |
| Portland      |   OR  |      603,106     |          1988         |
| San Jose      |   CA  |      982,765     |          2007         |
| Minneapolis   |   MN  |      392,880     |          2009         |
| New York      |   NY  |     8,199,221    |          2006         |

##Use vs. Process

Most cities focus on either restricting use directly (stricter zoning) or increasing process requirements (comp plan overlays). 


##Cities Use Combination of Restrictions and Process

```{r, echo=FALSE, message=FALSE, warning=FALSE}

land_matrix <- read_csv("policy_matrix.csv")

land_long <- gather(land_matrix, key = "Policy", value = "Restriction", 2:9)

land_long$Policy <- factor(land_long$Policy, levels = c("Residential", "Commercial", "Educational", 
                                                        "City.Review", "Region.State",
                                                        "Industrial.Impact", "CBA", "Industrial.Support"))

p1 <- ggplot(land_long, aes(x = City, y = Policy, fill = factor(Restriction))) + geom_tile() + scale_fill_manual(values = c("white", "blue" )) + coord_flip() + theme(axis.text.x  = element_text(angle = 45, hjust = 1))

p2 <- ggplot(land_long, aes(x = City, y = Policy, fill = Restriction)) + geom_tile(color = "white") + 
  scale_fill_gradient(low = "white", high = "steelblue") + coord_flip() + 
  theme(axis.text.x  = element_text(angle = 45, hjust = 1, face = "bold", size = 14),
        axis.text.y = element_text(face = "bold", size = 14),
        axis.title.x = element_text(size = 16),
        axis.title.y = element_text(size = 16),
        legend.position = "none") 

p2
```

##Baltimore Maritime Industrial Districts

> - New zoning code
> - Passed in 2004 and strongest of the policies
> - Outright forbids nearly any non-industrial marine uses 
> - Rezoning applications require support of surrounding land owners, the city planning department, state port authority must be informed and give guidance

##Los Angeles Employment Protection Districts

> - A comprehensive plan overlay
> - Passed in 2008 after extensive industrial land use research period
> - Staff directed to reject residential rezones and to develop supportive policy
> - Industrial impact statements are required and notice to surrounding property owners
> - In less intensive zones, industrial conversion requires community benefits including relocation assistance and job training

##New York Industrial Business Zones

> - Local community-economic development organizations oversee them
> - Primarily zones for subsidy delivery including: tax relocation credits, job training credits, and small business support provided by the IBZ supervisor
> - Verbal statement from the Mayor to prevent residential conversion and hotel encroachment
> - Ongoing battle to include real restrictions in the city zoning code

#Conclusions

##Cities Care About Industrial Land

- Nearly a quarter of the nation's largest cities have some form of industrial land preservation policy
- Most of these policies are *new* and in flux
- These policies are *variable* ranging from outright bans to no net-loss strategies to verbal committments from city power structures
- The policies are primarily regulatory in nature and do not proactively support industrial users
- Indicative of a burgeoning set of industrial land *strategies* and links job creation more directly land-use

##Next Steps

- To what extent are these policies effective in protecting (and even growing) industrial jobs?
- What is the role of the planner in cities where these protections are in place? Do we see adequate enforcement? Are these policies reflective of actual city priorities and goals? 


